---
title: "Final_625"
author: "Edward Shao"
date: "12/16/2022"
output: pdf_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(tidyverse)
library(splines)
library(MASS)
```

## Data loading
```{r}
d.625 <- read.csv("C:/Users/s-edw/Downloads/Conditions_Contributing_to_COVID-19_Deaths__by_State_and_Age__Provisional_2020-2022.csv")
d.625 <- d.625[d.625$Group == "By Month", ] # Examine the data monthly rather than by year or total
d.625 <- d.625[d.625$State != "United States", ] # Interested in each state individually, not the total for the US
d.625 <- d.625[(d.625$Age.Group != "All Ages") & (d.625$Age.Group != "Not stated"), ] # Remove aggregate and not stated groups for Age
d.625$totMonth <- (d.625$Year - 2020) * 12 + d.625$Month # Create numerical variable for number of months since Dec 2019 for model fitting
d.625 <- d.625[, -c(1:6, 10, 13, 14)] # Remove unnecessary columns (time cols, Group, ICD10 Codes, Flag, etc.)
```



## Data Cleaning
```{r}
d.625 <- d.625[d.625$Group == "By Month", ] # Examine the data monthly rather than by year or total
d.625 <- d.625[d.625$State != "United States", ] # Interested in each state individually, not the total for the US
d.625 <- d.625[(d.625$Age.Group != "All Ages") & (d.625$Age.Group != "Not stated"), ] # Remove aggregate and not stated groups for Age
d.625$totMonth <- (d.625$Year - 2020) * 12 + d.625$Month # Create numerical variable for number of months since Dec 2019 for model fitting
d.625 <- d.625[, -c(1:6, 10, 13, 14)] # Remove unnecessary columns (time cols, Group, ICD10 Codes, Flag, etc.)
```



###### POPULATION DATA FOR MAPPING RATES ######################################

```{r}
## Get population data for each state
d.pop <- read.csv("Population Data.csv")
d.pop <- d.pop[, -c(2,3)]
d.pop <- mutate(d.pop, across('Population.Estimate..July.1..2021..POP_2021.', str_replace, ',', ''))
d.pop <- mutate(d.pop, across('Population.Estimate..July.1..2021..POP_2021.', str_replace, ',', ''))
# Necessary to do twice or else it will only remove one comma, and the following command will not be able to coerce it to a numerical variable.
d.pop[, 2] <- as.numeric(d.pop[, 2])
```

## Data cleaning for population data
```{r}
## Manipulate columns to merge neatly with ggplot state data
colnames(d.pop) <- c("region", "Population")
d.pop$region <- tolower(d.pop$region)


```


## Visualization / Exploratory Analysis

### Basic Data visualization

## boxplot of conditions vs COVID-19 deaths factored by conditions.
```{r}
ggplot(subset(data,data$Condition%in%unique(data$Condition)[1:2]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[3:4]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[5:6]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[7:8]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[9:10]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[11:12]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[13:14]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[15:16]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[17:18]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[19:20]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[21:22]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))

ggplot(subset(data,data$Condition%in%unique(data$Condition)[23]), aes(x=as.factor(Condition), y=COVID.19.Deaths)) + geom_boxplot(aes(fill=as.factor(Age.Group)))
```








###### Mapping - Spatial analysis #############################################
## Mapping data

```{r}
states <- map_data("state")
d.625.mapping <- d.625[d.625$Age.Group == "All Ages", -c(1:6, 10, 13, 14)]
d.625.mapping <- aggregate(COVID.19.Deaths ~ totMonth + State, data = d.625.mapping, FUN = sum)
```



## Aggregate New York and New York City
```{r}

d.625.mapping$State[d.625.mapping$State == "New York City"] = "New York"
d.625.mapping <- aggregate(COVID.19.Deaths ~ State + totMonth, data = d.625.mapping, FUN = sum)

```


## Merge Datasets (population and states data)
```{r}
d.625.mapping$region <- tolower(d.625.mapping$State)

d.625.mapping <- merge(d.625.mapping, states, by = "region")
d.625.mapping <- merge(d.625.mapping, d.pop, by = "region")

```



## Create the COVID Deaths Per 100000 variable for plotting
```{r}
d.625.mapping$COVID.Deaths.Per.100000 <- d.625.mapping$COVID.19.Deaths / (d.625.mapping$Population / 100000)
```



## Print all maps for each time point - 2 second delay between maps
```{r}
for (i in 1:35) {
  print(ggplot(d.625.mapping[d.625.mapping$totMonth == i, ], aes(x = long, y = lat, group = group)) + geom_polygon(aes(fill = COVID.Deaths.Per.100000), color = "black") + scale_fill_gradient(low = "gray", high = "red", limits = c(0, 360)) + labs(title = paste("Number of Months:", as.character(i), sep = " "))) 
  Sys.sleep(2)
}

```



###### Time series for condition group - Temporal and conditions analysis #####
```{r}

## Subset the data for time series analysis

d.625.c <- d.625[d.625$Condition.Group != "COVID-19", ]

## Aggregate COVID Deaths over date and Condition group
d.625.c <- aggregate(COVID.19.Deaths ~ totMonth + Condition.Group, data = d.625.c, FUN = sum)

## Plot time series of COVID Deaths
ggplot(d.625.c, aes(x = totMonth, y = COVID.19.Deaths, group = Condition.Group)) + geom_line(aes(color = Condition.Group), size = 0.75) + xlab("Months Since Dec 2019") + ylab("Deaths from COVID-19")

```




## Modeling
###### IMPUTATION #############################################################
#### Impute each missing point from uniform(1,9)

```{r}
## Copy the dataset for imputation
d.625.imp <- d.625

## Impute from uniform(1,9)
set.seed(123456)
d.625.imp$COVID.19.Deaths[is.na(d.625.imp$COVID.19.Deaths)] = round(runif(length(d.625.imp$COVID.19.Deaths[is.na(d.625.imp$COVID.19.Deaths)]), 0.5, 9.49))

```


###### Regression Modeling ####################################################
```{r}
## Copy the dataset for modeling
d.625.m <- d.625.imp

## Aggregate New York and New York City
d.625.m$State[d.625.m$State == "New York City"] = "New York"
d.625.m <- aggregate(COVID.19.Deaths ~ State + Condition.Group + Condition + Age.Group + totMonth, data = d.625.m, FUN = sum)

## Remove COVID-19 deaths without co-existing conditions, as well as the All category
d.625.m <- d.625.m[(d.625.m$Condition.Group != "COVID-19") & (d.625.m$Condition.Group != "All other conditions and causes (residual)"), ]
```


###### Modelling several potential models #####################################
## Naive model - lm (Normal outcome) with linear time

```{r}
m.naive <- lm(COVID.19.Deaths ~ totMonth + Condition.Group + Age.Group + State, data = d.625.m)
summary(m.naive)
```

# Adjusted R^2 = 0.1269 - Fits well enough, but we know this model to not be very accurate

## lm with a linear spline
# Spline knots obtained from exploratory analysis with time series and condition groups,

```{r}
m.lm.spline <- lm(COVID.19.Deaths ~ bs(totMonth, degree = 1, knots = c(4, 9, 13, 18, 21, 23, 25)) + Condition.Group + Age.Group + State, data = d.625.m)
summary(m.lm.spline)
```

## glm with poisson link - linear spline
```{r}
m.glm.spline <- glm(COVID.19.Deaths ~ bs(totMonth, degree = 1, knots = c(4, 9, 13, 18, 21, 23, 25)) + Condition.Group + Age.Group + State, data = d.625.m, family = "poisson")
summary(m.glm.spline)
```



# Check for overdispersion? Negative Binomial model may be more approbriate
```{r}
pchisq(m.glm.spline$deviance, df=m.glm.spline$df.residual, lower.tail=FALSE)
```
# p-value = 0, reject the null - There is overdispersion

## glm with Negative Binomial link, linear spline for time
```{r}
m.nb.spline <- glm.nb(COVID.19.Deaths ~ bs(totMonth, degree = 1, knots = c(4, 9, 13, 18, 21, 23, 25)) + Condition.Group + Age.Group + State, data = d.625.m)
summary(m.nb.spline)
```

# Better fit - Dispersion parameter for Negative Binomial (0.6468) family taken to be 1

## Using GEE with negative binomial link, linear spline for time?
# m.nb.gee <- geem(COVID.19.Deaths ~ bs(totMonth, degree = 1, knots = c(4, 9, 13, 18, 21, 23, 25)) + Condition.Group + Age.Group, id = State, data = d.625.m, corstr = "exchangeable", family = negative.binomial(1))
# Unable to fit

## We can then consider the glm with a Negative Binomial link and a linear spline for time to be the final model.